Visualizing probability distributions across bivariate nonlinear temporal granularities

### 2020-02-24 main.Rmd
content cut - cyclic calendar algebra - iteration 1

**Discussion**: If linear granularity $G'$ is periodical with respect to $B$ with irregular mapping, then there exist $R'$, $P' \in \mathbb{Z}_+$ such that if $G'(w) = \bigcup_{z \in T_w}B(z)$ then $$G'(w) = \bigcup_{z \in T_w \mod R'}B(P'*\lfloor w/R' \rfloor + z)$$ (from definition \ref{def:periodical})

<!-- - $B \trianglelefteq G'$ -->

-

<!-- and,  number of granules of $B$ in each granule of $G'$ is not a constant (irregular mapping) -->

Here $w \mod R'$ represents the index that must be shifted to obtain $G'(w)$. The idea here is if we know the composition of each of the granules of $G'$ in terms of granules of $B$ for one period, we can find the composition of any granule of $G'$ beyond a period since the "pattern" repeats itself along the time domain due to the periodic property. The periodic property also ensures that $\vert T_w \vert = \vert T_{w\mod R'} \vert$ since every $w^{\text{th}}$ and $(w+R')^{\text{th}}$ granule of $G'$ will have the same number of granules of $B$. The term $\sum_{w=0}^{k-1}\vert T_{w}\vert$ denotes the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$. Since $\vert T_w \vert = \vert T_{w\mod R'} \vert$, the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$ becomes $\sum_{w=0}^{k-1}\vert T_{w \mod R'}\vert$ in Definition \autoref{def:quasicircular}.

@@

for $z \in T_k$ and $e_k$ is the number of exceptions in $\bigcup\limits_{w = 0}^{k} T_w$ and $E_{u} = [e_{begin}(u), e_{end}(u)]$.
## Aperiodic {#sec:aperiodic-gran-def}

@@

Aperiodic nonlinear granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$, that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity. Further assume that $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to \autoref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that
\begin{equation*}
M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)
\end{equation*}
and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{\text{th}}$ granule of the $i^{\text{th}}$ linear aperiodic granularity.

@@

The cyclic calendar algebra consists of broadly two kinds of operations:
(1) **single-to-multiple:** representation of multiple-order-up cyclic granularities from single order-up cyclic granularities
(2) **multiple-to-single:** representation of single-order-up cyclic granularities from multiple order-up cyclic granularities

The hierarchical structure of time creates a natural nested ordering which can produce **single-order-up** or **multiple-order-up** granularities. We shall use the notion of a hierarchy table and order to define them.

**Hierarchy table** Let $H_n: (G, C, P)$ be a hierarchy model containing n linear granularities. $G_{l}$ represents the linear granularity of order l and $P(l,m)$ represents the period length of $G_{l}$ with respect to $G_{m}$ and $C_{G(l),G(m)}$ represents the cyclic granularity that relates linear granularity of order l and m, $\forall l,m \in {1, 2, \dots,n}$ and $l<m$.

@@

We refer to granularities which are nested within multiple levels of the hierarchy table as multiple-order-up granularities and those concerning a single level as single-order-up granularities.

@@

<!-- some discussions on primary units of calendar -->
**Example:** So far we have used example of cyclic granularities from Gregorian calendar as it is the most widely used calendar. But it is far from being the only one. All calendars fall under three types - solar, lunar or lunisolar/solilunar but the day is the basic unit of time underlying all calendars [@Reingold2001-kf]. Various calendars, however, use different conventions to structure days into larger units: weeks, months, years and cycle of years. The French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds". Nevertheless, for any calendar a hierarchy can be defined. For example, in Mayan calendar, one day was referred to as 1 kin and the calendar was structured as follows:

 - 1 kin = 1 day
 - 1 uinal = 20 kin
 - 1 tun  = 18 uinal
 - 1 katun = 20 tun
 - 1 baktun = 20 katun

@@

*Example:* Let us use Equation \ref{eq:eq7} to compute the multiple-order-up granularity uinal_katun for Mayan calendar.

\begin{equation} \label{eq8}
\begin{split}
C_{uinal, baktun}(z)  = & L(C_{uinal, tun}(z) + P(uinal, tun)C_{tun,katun}(z) + C_{uinal, katun}C_{katun, baktun}(z)) \\
               = & L(\lfloor z/20\rfloor \mod 18  \\
               &~~~ + 20*\lfloor z/20*18\rfloor \mod 20  \\
               &~~~ + 20*18*20\lfloor z/20*18*20\rfloor \mod 20) \\
\end{split}
\end{equation}

@@

The multiple order-up cyclic granularites resulting from the first two cases are circular and has been handled in \autoref{sec:all-circular}. Any cyclic granularity resulting from the last case are quasi-circular. Examples might include hour-of-month or day-of-year. Let us consider the computation of $C_{hour, month}(z)$ for a $z$ such that $C_{hour, day}(z) = 12$, $C_{day, month}(z) = 20$ and $C_{month, year}(z) = 9$.

\begin{equation} \label{eq:hour-month}
\begin{split}
C_{hour, month}(z) & = L(C_{hour, day}(z) + P(hour, day)*C_{day, month}(z))\\
& = L(12 + 24*20)\\
\end{split}
\end{equation}

\begin{equation} \label{eq:day-year}
\begin{split}
C_{day, year}(z) & = L(C_{day,month}(z) + \sum_{w=0}^{C_{month, year}(z)-1}(\vert T_{w} \vert))\\
\end{split}
\end{equation}

where, $T_w$ is such that $month(w) = \bigcup_{z \in T_w}day(z)$ and $\vert T_w \vert$ is the cardinality of set $T_w$.

Clearly, the computation of multiple order-up granularities become more involved with quasi-circular granularities. $C_{hour, month}(z)$ can be represented as function of circular granularity $C_{hour, day}(z)$ and quasi-circular granularity $C_{day, month}(z)$, both of which are single order-up. However, $C_{day, year}(z)$ can not be computed only with single order-up granularities $C_{day, month}(z)$ and $C_{month, year}(z)$. It also requires the knowledge of the composition of the linear granularity $days$ within $months$ in an $year$ for computation of $\vert T_w \vert$.

Generally speaking, to keep it simple, we consider the case of only one quasi-circular single order-up granularity in the hierarchy. Any multiple order-up quasi-circular granularity can then be considered to be a function of some circular (single or multiple order-up) and a quasi-circular single order-up granularity.

<!-- Both of them can be circular or quasi-circular or it can be each of both. In case they are both quasi-circular or one circular and the other quasi-circular, we need to adjust the formula -  -->

For example, with $C_{l, m}(z) = f(C_{l,m'}(z), C_{m',m}(z))$, two different approaches need to be employed for computing multiple order-up granularities from single order-up ones.

- Case when $C_{l,m'}(z)$ is circular and $C_{m',m}(z)$ is quasi-circular

@@

- Case when  $C_{l,m'}(z)$  is quasi-circular and $C_{m',m}(z)$ is circular

@@

@@

If $C_{G_{l},G_{m'}}(z)$ or $C_{G_{l},G_{m'}}(z)$ are multiple order-up circular granularities, they could be computed using similar approach in Section \ref{sec:all-circular}.

>>>>

**Discussion**: If linear granularity $G'$ is periodical with respect to $B$ with irregular mapping, then there exist $R'$, $P' \in \mathbb{Z}_+$ such that if $G'(w) = \bigcup_{z \in T_w}B(z)$ then $$G'(w) = \bigcup_{z \in T_w \mod R'}B(P'*\lfloor w/R' \rfloor + z)$$ (from definition \ref{def:periodical})<!-- - $B \trianglelefteq G'$ --> <!-- and,  number of granules of $B$ in each granule of $G'$ is not a constant (irregular mapping) -->. Here $w \mod R'$ represents the index that must be shifted to obtain $G'(w)$. The idea here is if we know the composition of each of the granules of $G'$ in terms of granules of $B$ for one period, we can find the composition of any granule of $G'$ beyond a period since the "pattern" repeats itself along the time domain due to the periodic property. The periodic property also ensures that $\vert T_w \vert = \vert T_{w\mod R'} \vert$ since every $w^{\text{th}}$ and $(w+R')^{\text{th}}$ granule of $G'$ will have the same number of granules of $B$. The term $\sum_{w=0}^{k-1}\vert T_{w}\vert$ denotes the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$. Since $\vert T_w \vert = \vert T_{w\mod R'} \vert$, the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$ becomes $\sum_{w=0}^{k-1}\vert T_{w \mod R'}\vert$ in Definition \autoref{def:quasicircular}.

@@

for $z \in T_k$ and $e_k$ is the number of exceptions in $\bigcup\limits_{w = 0}^{k} T_w$ and $E_{u} = [e_{begin}(u), e_{end}(u)]$.

## Aperiodic {#sec:aperiodic-gran-def}

@@

Aperiodic nonlinear granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$, that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity. Further assume that $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to Definition \ref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$ and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{\text{th}}$ granule of the $i^{\text{th}}$ linear aperiodic granularity.

@@

The cyclic calendar algebra consists of broadly two kinds of operations:
(1) **single-to-multiple:** and (2) **multiple-to-single:** which entails the representation of multiple-order-up cyclic granularities from single order-up cyclic granularities and vice versa. The hierarchical structure of time creates a natural nested ordering which can produce _single-order-up_ or _multiple-order-up_ granularities. We shall use the notion of a hierarchy table and order to define them.

**Hierarchy table** Let $H_n: (G, C, P)$ be a hierarchy table containing $n$ linear granularities. $G_{l}$ represents the linear granularity of order l and $P(l,m)$ represents the period length of $G_{l}$ with respect to $G_{m}$ and $C_{G(l),G(m)}$ represents the cyclic granularity that relates linear granularity of order l and m, $\forall l,m \in {1, 2, \dots,n}$ and $l<m$.

@@

We refer to granularities which are nested within multiple levels of the hierarchy table as _multiple-order-up_ granularities and those concerning a single level as _single-order-up_ granularities.

@@

<!-- some discussions on primary units of calendar -->
**Example:** So far we have used example of cyclic granularities from Gregorian calendar as it is the most widely used calendar. But it is far from being the only one. <!--All calendars fall under three types - solar, lunar or lunisolar/solilunar-->. The day is the basic unit of time underlying most calendars [@Reingold2001-kf]. Various calendars, however, use different conventions to structure days into larger units: weeks, months, years and cycle of years. The French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds". Nevertheless, for any calendar a hierarchy table can be defined. For example, in Mayan calendar, one day was referred to as 1 kin and the calendar was structured such that 1 kin = 1 day; 1 uinal = 20 kin; 1 tun  = 18 uinal; 1 katun = 20 tun and 1 baktun = 20 katun

@@

**Example:** Let us use Equation \ref{eq:eq7} to compute the multiple-order-up granularity uinal_katun for Mayan calendar.

\begin{equation} \label{eq8}
\begin{split}
C_{uinal, baktun}(z)  = & L(C_{uinal, tun}(z) + P(uinal, tun)C_{tun,katun}(z) + C_{uinal, katun}C_{katun, baktun}(z)) \\
               = & L(\lfloor z/20\rfloor \mod 18 + 20*\lfloor z/20*18\rfloor \mod 20  \\
               &~~~ + 20*18*20\lfloor z/20*18*20\rfloor \mod 20) \\
\end{split}
\end{equation}

@@

The multiple order-up cyclic granularites resulting from the first two cases are circular and has been handled in \autoref{sec:all-circular}. Any cyclic granularity resulting from the last case are quasi-circular. Examples might include hour-of-month or day-of-year.

Generally speaking, to keep it simple, we consider the case of only one quasi-circular single order-up granularity in the hierarchy table. Any multiple order-up quasi-circular granularity $C_{l, m}(z)$ can then be considered to be a function of two single order-up granularities $C_{l,m'}(z)$ and $C_{m',m}(z)$. Two different approaches need to be employed for the follwing cases:

<!-- For example, with $C_{l, m}(z) = f(C_{l,m'}(z), C_{m',m}(z))$, -->
- $C_{l,m'}(z)$ is circular and $C_{m',m}(z)$ is quasi-circular

@@

- $C_{l,m'}(z)$ is quasi-circular and $C_{m',m}(z)$ is circular

@@

<!-- Let us consider the computation of $C_{hour, month}(z)$ for a $z$ such that $C_{hour, day}(z) = 12$, $C_{day, month}(z) = 20$ and $C_{month, year}(z) = 9$. -->

**Example:** Let us use Equations \ref{eq:multifromsingle-quasi1} and \ref{eq:multifromsingle-quasi2} to compute the multiple-order-up granularity hour_month and day_year for the Gregorian calendar considered.

 - $C_{hour, month}(z) = L(C_{hour, day}(z) + P(hour, day)*C_{day, month}(z))$  
 - $C_{day, year}(z) = L(C_{day,month}(z) + \sum_{w=0}^{C_{month, year}(z)-1}(\vert T_{w} \vert))$, where, $T_w$ is such that $month(w) = \bigcup_{z \in T_w}day(z)$.

<!-- CONTENT CUT -->
<!-- Clearly, the computation of multiple order-up granularities become more involved with quasi-circular granularities. $C_{hour, month}(z)$ can be represented as function of circular granularity $C_{hour, day}(z)$ and quasi-circular granularity $C_{day, month}(z)$, both of which are single order-up. However, $C_{day, year}(z)$ can not be computed only with single order-up granularities $C_{day, month}(z)$ and $C_{month, year}(z)$. It also requires the knowledge of the composition of the linear granularity $days$ within $months$ in an $year$ for computation of $\vert T_w \vert$. -->
<!-- If $C_{G_{l},G_{m'}}(z)$ or $C_{G_{l},G_{m'}}(z)$ are multiple order-up circular granularities, they could be computed using similar approach in Section \ref{sec:all-circular}. -->

@@
